Surface-driven registration method for the structure-informed segmentation of diffusion MR images\tnoteref{t1}

### 2016-03-02 Publications/2015-NeuroImage/04-methods.tex
remove comments from text

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}'\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:regseg-bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:regseg-transform} in the moving image.
For convenience and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions 
\revcomment[R\#3-C.11]{%
  $\mathcal{N} (\vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} )$}, with parameters
  $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$ \citep{esteban_mbis_2014}:

\revcomment[R\#3-C.15]{%
  \begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r}' \in \Omega_l}{\prod}
  \mathcal{N} (\vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} )
  \ifthenelse{\boolean{review}}{=}{= \notag\\ &=}
  \underset{l}{\prod} \underset{\vec{r}' \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:regseg-pdf}
  \end{align}%
  }%
  using $\mdist{f'}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}'$ with respect
  to the descriptors of region $l$ as
  $\mdist{f'}{l} = (\vec{f}' - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f}' - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
\todo{ordering} Even though the features being segmented are not generally i.i.d., the 
  spatial interdependency of voxels is implicitly supported by the piecewise smooth
  partition of the space \omegaset{}.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.
The \autoref{fig:regseg-model} shows \revcomment[authors]{how the joint distribution of 
  the input images} is approximated with a mixture of multivariate normal distributions,
  and this minimization is illustrated for the segmentation of the \gls*{fa} and the \gls*{adc} maps
  of one subject.

\begin{figure}
  \includegraphics[width=\linewidth]{figure01}
  \caption{The {\color{revcolor} interfacing surfaces $\Gamma_m$ between the competing \glspl{roi} $\Omega_l$,
  play the role of active contours which drive the registration process.}
  They evolve iteratively {\color{revcolor} along the normal $\hat{\mathbf{n}}_i$ of the surface} at each vertex
    $\vec{v}_i$ of the mesh.
  The gradient speeds $\bar{s}_i$ drive registration, which are computed as the disparity of the data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$.
  {\color{revcolor} The computation of shape-gradients is developed in \ref{app:shape_gradients}.}
  In this figure, the $\bar{s}_1$ derived from \autoref{eq:regseg-shape_gradient_disc2} is 
    written in the lower box, with $\Omega_{wm}$ in the inner limiting region, $\Omega_{gm}$
    the outer region, and $w_{0,1}$ is the relative area associated with vertex $\vec{v}_1$
    with respect to the total area of surface $\Gamma_0$.
  }\label{fig:regseg-method}
\end{figure}

\begin{figure*}
% \includestandalone[width=\linewidth]{figure02-model}
\begin{minipage}{\textwidth}
\includegraphics[width=\linewidth]{figure02}\\%
\vskip-2.0em
\includegraphics[width=\linewidth]{legend}%
\end{minipage}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$,
  for one real dataset.
{\color{revcolor}%
Panel (A, left) shows the joint distribution of the \gls*{fa} and \gls*{adc} conditioned to
  the segmentation \omegaset{} defined by the surfaces $\gammaset_R$ extracted from
  the \gls*{t1} image.
The plot was generated for reference using undistorted diffusion data, and
  therefore, $\gammaset_R$ is aligned with the \gls*{fa} and the \gls*{adc}.
The problem arises when the diffusion data present deformation, and the
  contours $\gammaset_R$ do not fit within the data (A, center).
After registration with \regseg{}, the contours are mapped onto the
  diffusion data (A, right), and the joint density plot is closer to the reference
  situation.
In panel (B), the three plots in (A) are decomposed tissue-wise.
Using filled contours, the bivariate distribution of each tissue is highlighted in its designated color,
  and represented over the remaining tissues (in gray colors).
To help assessment, dashed contours in black-to-white colors represent
  the corresponding distribution in the reference plot.
The registration process optimizes the segmentation model of \regseg{},
  and thus, the distribution of each region after registration is located
  closer to that corresponding in the reference situation, the shape of the distribution
  is more similar to the reference, and their spread is also reduced.
The effects of optimization are more noticeable on the \gls*{gm} ($\Omega_\text{GM}$) and
  the \gls*{wm} ($\Omega_\text{WM}$).
Particularly, the \gls*{wm} typically shows a bimodal distribution when the contours
  \gammaset{} do not fit the data.}
% The left plot in each panel represents a kernel density estimate of the location and spread of
%   each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
%   the \gls*{fa} and \gls*{adc} maps.
% At the top and the right margins, the marginal distributions of $\Omega_l$ on the \gls*{fa} and
%   \gls*{adc} are respectively displayed.
% On the right side, the sampled distribution of each region is shown in small scatter plot, in contrast to
%   the complete bivariate data.
% A) shows the initialization of the model, when the anatomically-correct surfaces are not
%   mapped into the \gls*{dmri} space.
% The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
%   \gls*{adc} maps with the original (undistorted) surfaces.
% Panel B shows the equivalent information, using the surfaces that are perfectly mapped into the
%   distorted \gls*{dmri} space.
% Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
%   reduced with respect to those in A.
% Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
%   are minimized with respect to panel A, and the resulting segmentation is close
%   to the theoretical situation presented in panel B.}
}\label{fig:regseg-model}
\end{figure*}

\paragraph*{Regularization}
The smoothness of the resulting \revcomment[R\#3-C.33]{displacement field} is induced
  by a Thikonov regularization prior:

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:regseg-thikonov}
  \end{align}
 which requires that the distortion and its gradient have zero mean\revcomment[R\#1-C4]{,}
   and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
\revcomment[R\#3-C.16]{%
Therefore, $\mathbf{A}$ and $\mathbf{B}$ are tensors that modulate the regularization, and
  produce deformations with preferential directions}.
Finally, the maximum a posteriori problem is adapted to a variational problem where we search for
  the minimum of an energy functional by applying $E(\vec{u}) = -\log \{P( M \mid U) \, P(U)\}$:

\revcomment[R\#3\\C.17--C.19]{%
  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r}' \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) =
\notag\\ &= - \underset{l}{\sum} \int_{\Omega_l} \Big\{ \log \left[ \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right) \right] \, + \log \left[ p_0( \vec{u})\,p_1( \vec{u}) \right] \Big\} \, d\vec{r}'
   = \notag\\ &=
  \const + \underset{l}{\sum} \Big\{ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \Big\} \,
  + \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}'.
  \label{eq:regseg-energy}
  \end{align}%
}
This expression is the dual of the Mumford-Shah functional that corresponds
  to the framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:regseg-general_gradient_descent}
  \end{equation}
  where $t$ is an artificial time parameter of the contour
  evolution and $\vec{u}_k$ are the parameters that support the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that \revcomment[R\#1-C.5\\R\#3-C.31]{the preferential directions of
  the displacement are aligned} with the imaging axes to simplify \eqref{eq:regseg-energy} as expression 
  \eqref{eq:regseg-app_energy} in \ref{app:reg_term}, and thus to compute its
  derivative \eqref{eq:regseg-general_gradient_descent}:

\revcomment[R\#3-C.20]{%
  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \big[ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r}' \big] \,
  + \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}'
  \Big\},
  \label{eq:regseg-gradient_descent}
  \end{align}}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$, \revcomment[R\#3-C.21]{%
  and $\{\boldsymbol{\alpha}, \boldsymbol{\beta}\}$ are the expected variances along
  the imaging axes of the displacement field and its gradient, respectively.}
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
\revcomment[R\#3-C.14]{%
The derivative of the data term is computed using explicit shape gradients (see \ref{app:shape_gradients}),
  which finally lead to obtain vertex-wise speeds of the gradient $\bar{s}_i$ as}
  illustrated in \autoref{fig:regseg-method}.
\revcomment[R\#3-C.14]{%
Using the expressions \eqref{eq:regseg-gradient_wshape} and 
  \eqref{eq:regseg-shape_gradient_final} given in \ref{app:shape_gradients}
  and introducing the analytical derivative of the regularization term, then
  the \autoref{eq:regseg-gradient_descent} is reformulated as:}

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \vec{g}_k  + \boldsymbol{\alpha} \cdot \vec{u}_k - \boldsymbol{\beta} \cdot (\Delta \vec{u}_k),
  \label{eq:regseg-final_gradient}
  \end{align}
\revcomment[R\#3-C22]{%
where $\vec{g}_k$ are the shape-gradient contributions on the coefficients $\vec{u}_k$ of the B-spline grid.}
Finally, to descend this gradient, we establish a semi-implicit Euler scheme
  \citepalias[see][section S1.3]{esteban_useful_2015},
  with a step size parameter $\delta$, which we solve in the spectral domain as follows:

@@

\paragraph*{Implementation details and convergence}%
\label{sec:regseg-conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between the control points of the B-spline basis
  functions for each level of the multiresolution strategy.
\revcomment[R\#3-C32]{As a rule of thumb, for a $\delta =$ 1.0, both $\boldsymbol{\alpha}$
  and $\boldsymbol{\beta}$ will typically be in the range [0.0, 1.0].
The parameters used ($\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-spline grid resolutions,
  and target image smoothing), the implementation details, and other features such as the sparse matrix
  approach to fast interpolation} are discussed in the \citetalias[][section S1]{esteban_useful_2015}.

@@

A first proof of concept is introduced to demonstrate \regseg{} in digital phantoms
  with simple geometries, using $U_\text{true}$ without directional restrictions.
Then, \regseg{} is evaluated in a framework using undistorted \gls*{dmri} datasets,
  and $U_\text{true}$ is derived from the corresponding inhomogeneity fieldmap of
  the subject.
Therefore, the deformation field is nonzero only in the \revcomment[R\#3-C.23]{\acrfull*{pe}} axis,
  and \revcomment[R\#1-C3]{reproduces a} real \gls*{epi} distortion.
The adaptation of the evaluation protocol to the simulated phantoms and the real data is
  explained in the following sections.

@@

  \begin{equation}
  sWI = \frac{1}{\sum_i a_i} \sum\limits_i^{N_V} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|,
  \label{eq:regseg-swindex}
  \end{equation}
  where $\vec{v}_i \subset \gammaset_\text{true}$ are the locations of the total $N_V$ vertices,
  \revcomment[R\#3-C.24]{$a_i$ is the area corresponding to each vertex $\vec{v}_i$},
  and $\hat{\vec{v}}_i \subset \hat{\gammaset}_\text{test}$ are the recovered locations
  that correspond to $\vec{v}_i$.
In practice, we only report the \gls*{swindex} for three surfaces (\{$\Gamma_{VdGM}, \Gamma_{WM}, \Gamma_{pial}$\})
  of crucial interest in whole-brain tractography.

>>>>

  \begin{equation}
  P(M \mid U) = \underset{l}{\prod} \underset{\vec{r}'\in \Omega_l}{\prod}
    P\left( \vec{f}' \mid U \right),
  \label{eq:regseg-bayes_aposteriori}
  \end{equation}
  where $\vec{f}' = M(\vec{r}')$ is the feature vector at the displaced
  position $\vec{r}'$ \eqref{eq:regseg-transform} in the moving image.
For convenience and because it has been shown to be an appropriate approximation
  \citep{leemput_automated_1999,cuadra_comparison_2005}, we introduce two assumptions for each
  region $\Omega_l$:
  1) the features are i.i.d.; and 2) they can be modeled by multivariate normal
  distributions 
  $\mathcal{N} (\vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} )$, with parameters
  $\lbrace \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} \rbrace$
  for each region $\Omega_l$ \citep{esteban_mbis_2014}:

  \begin{align}
  P( M \mid U) &= \underset{l}{\prod} \underset{\vec{r}' \in \Omega_l}{\prod}
  \mathcal{N} (\vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_{l} )
  \ifthenelse{\boolean{review}}{=}{= \notag\\ &=}
  \underset{l}{\prod} \underset{\vec{r}' \in \Omega_l}{\prod} \frac{1}{ \sqrt{(2\pi)^{C}\,\left|\boldsymbol{\Sigma}_{l}\right|}}\,{e^{\left(-\frac{1}{2}
  \mdist{f'}{l} \right)}},
  \label{eq:regseg-pdf}
  \end{align}%
  using $\mdist{f'}{l}$ to denote the squared \emph{Mahalanobis distance} of $\vec{f}'$ with respect
  to the descriptors of region $l$ as
  $\mdist{f'}{l} = (\vec{f}' - \boldsymbol{\mu}_l)^T \, {\boldsymbol{\Sigma}_l}^{-1} \, (\vec{f}' - \boldsymbol{\mu}_l)$.
$C$ is the number of channels comprised in the image $M$.
Even though the features being segmented are not generally i.i.d., the 
  spatial interdependency of voxels is implicitly supported by the piecewise smooth
  partition of the space \omegaset{}.
In fact, the projection of $\gammaset_R$ onto $M$ is an implicit segmentation model, for which
  the covariance matrix $\boldsymbol{\Sigma}_l$ of each region is minimized.
The \autoref{fig:regseg-model} shows how the joint distribution of 
  the input images is approximated with a mixture of multivariate normal distributions,
  and this minimization is illustrated for the segmentation of the \gls*{fa} and the \gls*{adc} maps
  of one subject.

\begin{figure}
  \includegraphics[width=\linewidth]{figure01}
  \caption{The interfacing surfaces $\Gamma_m$ between the competing \glspl{roi} $\Omega_l$,
  play the role of active contours which drive the registration process.
  They evolve iteratively along the normal $\hat{\mathbf{n}}_i$ of the surface at each vertex
    $\vec{v}_i$ of the mesh.
  The gradient speeds $\bar{s}_i$ drive registration, which are computed as the disparity of the data
    energies with respect to the two limiting regions of $M(\vec{v}_i)$, the features of the image
    $M$ in the location of vertex $\vec{v}_i$.
  The computation of shape-gradients is developed in \ref{app:shape_gradients}.
  In this figure, the $\bar{s}_1$ derived from \autoref{eq:regseg-shape_gradient_disc2} is 
    written in the lower box, with $\Omega_{wm}$ in the inner limiting region, $\Omega_{gm}$
    the outer region, and $w_{0,1}$ is the relative area associated with vertex $\vec{v}_1$
    with respect to the total area of surface $\Gamma_0$.
  }\label{fig:regseg-method}
\end{figure}

\begin{figure*}
% \includestandalone[width=\linewidth]{figure02-model}
\begin{minipage}{\textwidth}
\includegraphics[width=\linewidth]{figure02}\\%
\vskip-2.0em
\includegraphics[width=\linewidth]{legend}%
\end{minipage}
% \includegraphics[width=\linewidth]{figure04B}
% \includegraphics[width=\linewidth]{figure04C}
\caption{%
Evolution of the segmentation model defined by the homogeneous regions $\Omega_l$,
  for one real dataset.
Panel (A, left) shows the joint distribution of the \gls*{fa} and \gls*{adc} conditioned to
  the segmentation \omegaset{} defined by the surfaces $\gammaset_R$ extracted from
  the \gls*{t1} image.
The plot was generated for reference using undistorted diffusion data, and
  therefore, $\gammaset_R$ is aligned with the \gls*{fa} and the \gls*{adc}.
The problem arises when the diffusion data present deformation, and the
  contours $\gammaset_R$ do not fit within the data (A, center).
After registration with \regseg{}, the contours are mapped onto the
  diffusion data (A, right), and the joint density plot is closer to the reference
  situation.
In panel (B), the three plots in (A) are decomposed tissue-wise.
Using filled contours, the bivariate distribution of each tissue is highlighted in its designated color,
  and represented over the remaining tissues (in gray colors).
To help assessment, dashed contours in black-to-white colors represent
  the corresponding distribution in the reference plot.
The registration process optimizes the segmentation model of \regseg{},
  and thus, the distribution of each region after registration is located
  closer to that corresponding in the reference situation, the shape of the distribution
  is more similar to the reference, and their spread is also reduced.
The effects of optimization are more noticeable on the \gls*{gm} ($\Omega_\text{GM}$) and
  the \gls*{wm} ($\Omega_\text{WM}$).
Particularly, the \gls*{wm} typically shows a bimodal distribution when the contours
  \gammaset{} do not fit the data.
% The left plot in each panel represents a kernel density estimate of the location and spread of
%   each region $\Omega_l$ in the bivariate feature space of $M$, which comprises
%   the \gls*{fa} and \gls*{adc} maps.
% At the top and the right margins, the marginal distributions of $\Omega_l$ on the \gls*{fa} and
%   \gls*{adc} are respectively displayed.
% On the right side, the sampled distribution of each region is shown in small scatter plot, in contrast to
%   the complete bivariate data.
% A) shows the initialization of the model, when the anatomically-correct surfaces are not
%   mapped into the \gls*{dmri} space.
% The panel shows the resulting conditional distributions of segmenting the (distorted) \gls*{fa} and
%   \gls*{adc} maps with the original (undistorted) surfaces.
% Panel B shows the equivalent information, using the surfaces that are perfectly mapped into the
%   distorted \gls*{dmri} space.
% Since the surfaces are now aligned in $M$, the spreads of the conditional distributions for each region are
%   reduced with respect to those in A.
% Panel C, after \regseg{} registration, shows how the covariance matrices $\boldsymbol{\Sigma}_l$
%   are minimized with respect to panel A, and the resulting segmentation is close
%   to the theoretical situation presented in panel B.}
}\label{fig:regseg-model}
\end{figure*}

\paragraph*{Regularization}
The smoothness of the resulting displacement field is induced
  by a Thikonov regularization prior:

  \begin{align}
  P(U) = \underset{\vec{r}}{\prod}\, p(\vec{u}) &=
  \underset{\vec{r}}{\prod}\, p_0(\vec{u}) \, p_1(\vec{u}), \text{ with} \\
  p_0(\vec{u}) &= \mathcal{N}( \vec{u} \mid 0, \mathbf{A}^{-1}), \notag\\
  p_1(\vec{u}) &= \mathcal{N}(  \nabla \vec{u} \mid 0, \mathbf{B}^{-1}),
  \label{eq:regseg-thikonov}
  \end{align}
 which requires that the distortion and its gradient have zero mean,
   and variance governed by the matrices $\mathbf{A}$ and $\mathbf{B}$.
Therefore, $\mathbf{A}$ and $\mathbf{B}$ are tensors that modulate the regularization, and
  produce deformations with preferential directions.
Finally, the maximum a posteriori problem is adapted to a variational problem where we search for
  the minimum of an energy functional by applying $E(\vec{u}) = -\log \{P( M \mid U) \, P(U)\}$:

  \begin{align}
  E(\vec{u}) &= -\log \underset{l}{\prod}
  \underset{\vec{r}' \in \Omega_l}{\prod}
  \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right)\,p_0( \vec{u})\,p_1( \vec{u}) =
\notag\\ &= - \underset{l}{\sum} \int_{\Omega_l} \Big\{ \log \left[ \mathcal{N} \left( \vec{f}' \mid \boldsymbol{\mu}_l, \boldsymbol{\Sigma}_l \right) \right] \, + \log \left[ p_0( \vec{u})\,p_1( \vec{u}) \right] \Big\} \, d\vec{r}'
   = \notag\\ &=
  \const + \underset{l}{\sum} \Big\{ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r} \Big\} \,
  + \int_{\Omega} \frac12 \left[ \vec{u}^T \mathbf{A} \vec{u} + (\nabla \vec{u})^T \mathbf{B} (\nabla \vec{u}) \right] \,d\vec{r}'.
  \label{eq:regseg-energy}
  \end{align}%
This expression is the dual of the Mumford-Shah functional that corresponds
  to the framework of \acrlong*{acwe} \citep{chan_active_2001}
  with the anisotropic regularization term of \cite{nagel_investigation_1986}.

@@

  \begin{equation}
  \frac{\partial \vec{u}(\vec{r},t)}{\partial t} \propto - \frac{\partial E(\vec{u})}{\partial \vec{u}_k},
  \label{eq:regseg-general_gradient_descent}
  \end{equation}
  where $t$ is an artificial time parameter of the contour
  evolution and $\vec{u}_k$ are the parameters that support the estimate
  $\hat{U}$ of the transformation at the current time point.
Let us assume that the preferential directions of
  the displacement are aligned with the imaging axes to simplify \eqref{eq:regseg-energy} as expression 
  \eqref{eq:regseg-app_energy} in \ref{app:reg_term}, and thus to compute its
  derivative \eqref{eq:regseg-general_gradient_descent}:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \frac{ \partial }{\partial \vec{u}_k} \Big\{
  \underset{l}{\sum} \big[ \int_{\Omega_l} \mdist{f'}{l} \,d\vec{r}' \big] \,
  + \int_{\Omega} \frac12 [ \boldsymbol{\alpha} \cdot \vec{u}^{\circ2}
  + \boldsymbol{\beta} \cdot (\nabla \vec{u})^{\circ2} ] \,d\vec{r}'
  \Big\},
  \label{eq:regseg-gradient_descent}
  \end{align}
  where $\vec{u}^{\circ2} = \vec{u}^T \cdot \vec{u}$,
  and $\{\boldsymbol{\alpha}, \boldsymbol{\beta}\}$ are the expected variances along
  the imaging axes of the displacement field and its gradient, respectively.
Then, the data and regularization terms are split and discretized to compute their
  derivatives.
The derivative of the data term is computed using explicit shape gradients (see \ref{app:shape_gradients}),
  which finally lead to obtain vertex-wise speeds of the gradient $\bar{s}_i$ as
  illustrated in \autoref{fig:regseg-method}.
Using the expressions \eqref{eq:regseg-gradient_wshape} and 
  \eqref{eq:regseg-shape_gradient_final} given in \ref{app:shape_gradients}
  and introducing the analytical derivative of the regularization term, then
  the \autoref{eq:regseg-gradient_descent} is reformulated as:

  \begin{align}
  \frac{\partial E(\vec{u})}{\partial \vec{u}_k} &=
  \vec{g}_k  + \boldsymbol{\alpha} \cdot \vec{u}_k - \boldsymbol{\beta} \cdot (\Delta \vec{u}_k),
  \label{eq:regseg-final_gradient}
  \end{align}
where $\vec{g}_k$ are the shape-gradient contributions on the coefficients $\vec{u}_k$ of the B-spline grid.
Finally, to descend this gradient, we establish a semi-implicit Euler scheme
  \citepalias[see][section S1.3]{esteban_useful_2015},
  with a step size parameter $\delta$, which we solve in the spectral domain as follows:

@@

\paragraph*{Implementation details and convergence}%
\label{sec:regseg-conv_report}
The \regseg{} tool includes a multiresolution strategy on the free-form deformation field.
Registration pyramids are created by setting the spacing between the control points of the B-spline basis
  functions for each level of the multiresolution strategy.
As a rule of thumb, for a $\delta =$ 1.0, both $\boldsymbol{\alpha}$
  and $\boldsymbol{\beta}$ will typically be in the range [0.0, 1.0].
The parameters used ($\delta$, $\boldsymbol{\alpha}$, $\boldsymbol{\beta}$, the B-spline grid resolutions,
  and target image smoothing), the implementation details, and other features such as the sparse matrix
  approach to fast interpolation are discussed in the \citetalias[][section S1]{esteban_useful_2015}.

@@

A first proof of concept is introduced to demonstrate \regseg{} in digital phantoms
  with simple geometries, using $U_\text{true}$ without directional restrictions.
Then, \regseg{} is evaluated in a framework using undistorted \gls*{dmri} datasets,
  and $U_\text{true}$ is derived from the corresponding inhomogeneity fieldmap of
  the subject.
Therefore, the deformation field is nonzero only in the \acrfull*{pe} axis,
  and reproduces a real \gls*{epi} distortion.
The adaptation of the evaluation protocol to the simulated phantoms and the real data is
  explained in the following sections.

@@

  \begin{equation}
  sWI = \frac{1}{\sum_i a_i} \sum\limits_i^{N_V} a_i\,\|
  \vec{v}_i - \hat{\vec{v}}_i \|,
  \label{eq:regseg-swindex}
  \end{equation}
  where $\vec{v}_i \subset \gammaset_\text{true}$ are the locations of the total $N_V$ vertices,
  $a_i$ is the area corresponding to each vertex $\vec{v}_i$,
  and $\hat{\vec{v}}_i \subset \hat{\gammaset}_\text{test}$ are the recovered locations
  that correspond to $\vec{v}_i$.
In practice, we only report the \gls*{swindex} for three surfaces (\{$\Gamma_{VdGM}, \Gamma_{WM}, \Gamma_{pial}$\})
  of crucial interest in whole-brain tractography.
